---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# dtdatasources: Data Sources for Shiny Datatables

<!-- badges: start -->
<!-- badges: end -->

## Overview

`dtdatasources` provides Shiny [server-side datatables](https://rstudio.github.io/DT/server.html) backends for the DT package, allowing you to connect your datatables directly to databases or APIs, in addition
to the standard dataframes.


## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("grahamrp/dtdatasources")
```
## Available Data Sources

* SQLite
  - [x] paging
  - [x] sorting
  - [ ] filtering
  - [ ] row selection

So far I've only written a basic SQLite implementation, but would like to create
more. If you would like to help, please see the [Contributing](#contributing) section.

## About Custom DT Data Sources

When you want to use `DT::renderDataTables`/`DT::renderDT` with large datasets you can choose to render the table on the server with `renderDT(big_dataset, server = TRUE)`. This will perform the dataset filtering/sorting/paging on the server where Shiny
is running, instead of in the user's web browser, making the table more responsive.

The default DT implementation of server-side processing still requires the data
to be in a *dataframe* on the server. For very large datasets, or where the data
is stored outside the Shiny process, for example in a database or API, we would
not want to copy the entire dataset into a dataframe on the server, as it may
not have enough resources to hold it in memory, or enough processing power to
perform the server-side filtering/sorting/paging.

For this reason, `renderDT` provides a `funcFilter` parameter to supply
our own function that describes how to fetch, filter, sort, and page our dataset,
regardless of where the data resides.

The `dtdatasources` package will provide implementations of `funcFilter` for
various datasources, for you to use directly, or as examples to adapt to your
own implementations.


## Sqlite Example

This example shows how to connect a server-side datatable to a table in a SQLite database.
Take a look at [inst/app.R](https://github.com/grahamrp/dtdatasources/blob/master/inst/app.R) for the shiny app, or run `shiny::runApp(system.file("app.R", package = "dtdatasources"))`.

```{r example, eval=FALSE}
library(shiny)
library(DT)
library(DBI)
library(dtdatasources)

# Setup an example database and table
con <- dbConnect(RSQLite::SQLite(), ":memory:")
dbWriteTable(con, "mtcars", mtcars)

ui <- fluidPage(
  DTOutput("tbl")
)

server <- function(input, output, session) {

  # Create an initial dataframe. This only needs to contain column names, and
  # need not have any rows.
  initial_df <- dbGetQuery(con, "SELECT * FROM mtcars LIMIT 0;")

  # Create a funcFilter function describing how to get data for a datatable.
  # The filter factory 
  mtcars_filter <- dtdatasources::sql_filter_factory(
    # Use the sqlite connection created above
    con = con,
    # Set the query_fun to query_sqlite, or implement your own version
    query_fun = dtdatasources::query_sqlite,
    # Any additional args are passed into query_sqlite. In query_sqlite's case
    # it accepts a `tbl` parameter for which table to use.
    tbl = "mtcars"  
  )

  # Call renderDT with our custom filter function
  output$tbl <- renderDT(
    initial_df,
    server = TRUE,  # Must be TRUE to perform processing in R, not in the browser
    rownames = FALSE,  # Must be FALSE for the query_sqlite implementation
    funcFilter = mtcars_filter  # Provide the sqlite function filter created above
  )
}

shinyApp(ui, server)
```

## Making New Data Sources

If you want to use this example to create your own implementations of DT data
sources, take a look at `sqlite.R` and provide your own version of the
`query_sqlite` function for different databases. This function adapts the
`DT:::dataTablesFilter` code on [GitHub](https://github.com/rstudio/DT/blob/master/R/shiny.R), 
which is the default `funcFilter` for filtering/sorting/paging dataframes on the server.

## Contributing

Feedback, bug reports, fixes, and feature requests are welcome; file issues or seek support [here](http://github.com/grahamrp/dtdatasources/issues).

## Some References for `funcFilter`

| Description | URL |
| ------------ | --- |
| Description of the overall problem | [Google Groups shiny-discuss](https://groups.google.com/forum/#!msg/shiny-discuss/zaPqkMdhwy4/jHGFwBfEBQAJ)
| Open issue to make a funcFilter example  | https://github.com/rstudio/DT/issues/194
| Custom filtering problem with filter ranges and `filterRow()` | https://github.com/rstudio/DT/issues/50
| Row selection | https://github.com/rstudio/DT/issues/75
